>

### 2019-10-04 manuscript/manuscript.Rmd
NEW DRAFT

---
title: >
  Modeling the impacts of park access on health outcomes: a utility-based 
  accessibility approach
author:
  - name: Gregory Macfarlane
    email: gregmacfarlane@byu.edu
    affiliation: Brigham Young University
    corresponding: gregmacfarlane@byu.edu
  - name: Nico Boyd
    email:
    affiliation: Fehr and Peers 
  - name: John E. Taylor
    email: jetaylor@gatech.edu
    affiliation: Georgia Institute of Technology
  - name: Kari Watkins
    email: kariwatkins@gatech.edu
    affiliation: Georgia Institute of Technology
address:
  - code: Brigham Young University
    address: Civil and Environmental Engineering Department, Brigham Young University 430 Engineering Building, Provo, UT, 84602
  - code: Fehr and Peers
    address: Los Angeles, California
  - code: Georgia Institute of Technology
    address: School of Civil and Environmental Engineering, Georgia Institute of Technology, 755 Atlantic Drive, Atlanta, GA, 30332
abstract: >
  Recent research has underscored the potential for public green spaces to
  influence individual and societal health outcomes, but empirical measurements
  of such influences have yielded mixed results to date, with particular
  disagreement surrounding how access to parks ought to be defined while
  controlling alternate explanations. In this paper we present a comprehensive
  measure of park accessibility drawn from random utility choice theory, which
  avoids arbitrary assertions of proximity while incorporating potentially
  numerous amenities and attributes of both the parks and the population. We
  apply this choice-based accessibility measure to correlate Census tract-level
  obesity and physical activity rate estimates from the Centers for Disease
  Control and Prevention's 500 Cities project with tract-level American
  Community Survey socioeconomic data in New York City, paired with geographic
  parks data from New York City. Controlling for the
  socioeconomic variables and spatially correlated error terms, we show a
  positive and significant relationship between park access and physical
  activity rates, and a clear and significant negative relationship between park
  access and obesity rates. In doing so, this research contributes a more
  comprehensive modeling approach for measuring the impact of park access on
  health, and may improve our understanding of the role  parks and access to
  them can serve in furthering public health objectives.
author_summary: >
  Greg Macfarlane is an assistant professor of Civil Engineering at Brigham
  Young University in Provo, Utah.

@@

knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",", drop0trailing = TRUE)
})

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
Over the last several years a new social model of health has evolved, treating health
``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences.''[@Duhl1999, p.7]
In this new paradigm, resources provided via civil infrastructure --- in
particular parks and other public green spaces --- play a crucial role in
promoting and sustaining public health [@Bedimo-Rung2005; Wells2007; @Coutts2008].
After all, it does little to lecture individuals on the importance of exercise
if their community is built in such a way that such exercise is expensive,
unenjoyable, or unsafe.

In spite of the potential for green spaces to influence public health outcomes 
and numerous studies attempting to empirically estimate the strength of these
outcomes, evidence to date has been mixed. A major challenge researchers have
faced is a proliferation of techniques for measuring access to green spaces, of
which many are plausible but lack robust theoretical basis. In this
paper, we present a holistic and flexible measurement for park accessibility
based in activity location choice theory. This utility-based accessibility measure
compares the continuous distance to all parks in the region, weighted against
the sizes of each park and its other assorted amenities.

We apply this measurement to study the link between
park accessibility and attractiveness and Census tract-level aggregate physical
activity participation and obesity rates in New York City, controlling for
spatially correlated unobserved effect and tract-level socioeconomic
characteristics. We find a positive relationship where the least park-accessible
tracts have an expected physical activity participation rate roughly a full
percentage point lower than the most accessible tracts; a traditional
quarter-mile buffer analysis estimates a similar albeit marginally less
significant effect. However, we also find a strong and significant correlation
between the choice-based park accessibility metric and decreased obesity rates
--- again controlling for spatial correlation, socioeconomic attributes, and
physical activity rates. In this case a traditional buffer analysis does not
show significance.

We also demonstrate how the utility-based accessibility metric can potentially be
expanded to account for other measures of a park's attractiveness including its
observed social media activity and the presence of various park amenities.
The paper concludes with a discussion of opportunities for future research.

@@

The large majority of extant research on the topic of park access and public health
outcomes has been involved comparing metropolitan regions against each other. For example,
a study by @West2012 used park data from
the Trust for Public Land’s 2010 City Park Facts and public health data from the
Behavioral Risk Factor Surveillance System (BRFSS) to examine the relationship
between the density of park land and physical activity and obesity rates for 67
metropolitan statistical areas in the U.S. The findings in this case conformed to 
expectation; that is, the study found a significant positive association
between park density and physical activity and a significant negative
association between park density and obesity.
@Larson2016 similarly used self-reported scores on the Gallup-Healthways
Wellbeing Index to evaluate the relationship between physical health and park
quantity, quality, and accessibility in 44 U.S. cities. While the authors found
positive associations between wellbeing and park quality and accessibility,
these relationships were not statistically significant.
Conversely, @Richardson2012 examined the relationship between total urban
greenspace and mortality rates for selected maladies; though the authors did not
find a statistically significant relationship between the quantity of urban
greenspace and mortality caused by lung cancer, diabetes, heart disease or car
accidents, they did find that all-cause mortality was oddly, *higher* in greener
cities. Finally, in a meta-analysis of 20 peer reviewed journal articles exploring the
relationship between parks and objectively measured physical activity,
@Bancroft2015 found that five studies reported a significant positive
association between the two, six studies produced mixed results, and nine
studies found no association at all.

@@

Loosely defined, the *accessibility* of an arbitrary place describes the ease
with which people can accomplish activities there. Accessibility is an abstract
concept with tempting quantifiability [@Handy1997], perhaps explaining the proliferation of
calculation mechanisms. @Dong2006 present a helpful mathematical heirarchy of
common mechanisms, which we follow here. Consider a person residing at
point $i$ in a city with parks $j \in 1 \ldots J$. An analyst might consider
point $i$ as "having access" to a park if the distance to the park $d_ij$  within an
"isochrone," or less than some threshold $D$,

@@

Variations of this isochrone-based framework are easily derived and relatively
common. 
@Dias2019 considered road safety perception as an extenuating factor in the
effective distance to a park.
ParkScore [@parkscore2019] is a sophisticated application of this approach where $d_{ij}$ is a
carefully calculated 10-minute network-bound walk, with a demographic analysis of
areas within and without this threshold. In other applications $D$ might be more
variably defined, such as the presence or percent of green space within a
political or statistical boundary [@Mitchell2008].
The indicator function could also be modified into the total number of parks or
percent of park space within radius $D$ of point $i$ [@Kascynski2014].
In a study of neighborhoods in New York
City, @Stark2014 found that a higher proportion of park space in a neighborhood
was associated with a lower BMI for its residents, and lower park cleanliness
scores were associated with a higher BMI, suggesting that increasing the supply
of clean, well-maintained parks could positively impact adult BMI.

In spite of the flexibility of adapted isochrone techniques, the arbritrary
definitions necessarily imposed by researchers in its application can inhibit
holistic analysis [@Logan2017].
A somewhat more complete approach is the so-called "gravity" accessibility
statistic. In this case the accessibility of a point is the denominator of the
gravity formulation of a trip distribution model,
\begin{equation}\label{eq:gravity}
A_i = \sum_{j = 1}^J S_j f(d_{ij})
\end{equation}
where $S_j$ is the "size" of each park literally (i.e. acres) or abstractly (i.e.
trip attractions) and $f(d_ij)$ is a monotonically decreasing cost function,
usually a negative exponential. @Dong2006 note that the isochrone framework is a
special case of the gravity model where $f(d_{ij})$ is a binary function. 
@Giles-Corti2005 compare various gravity-based accessibility scores with an
isochrone specification and show that the former are more predictive of walking
behavior; that is, the attributes of a park matter influence walking more than
merely the distance to it. @Zhang2011 present a population-weighted
gravity-based accessibility to parks metric with a national rather than
metropolitan scope, though they do not examine its correlation with health
outcomes.

A third accessibility mechanism is a utility-based specification, termed as such
for being derived from random utility choice theory. Consider that an individual
living at point $i$ is choosing a park for a recreation activity. If we apply
the multinomial logit model [@McFadden1974], the expected consumer surplus
enjoyed by this individual can be shown to be
\begin{equation}\label{eq:utility}
A_i = \ln\left({\sum_{j = 1}^J\exp(V_{ij})}\right) + C
\end{equation}
where parks are differentiated from each other by their relative measurable
*utilities* $V_{ij}$. $C$ is an unknown constant, but the difference in consumer
surplus between two points $i$ and $k$ can be quantitatively compared as 
$A_i - A_k$ [@Bruce1977]. In principle, $V$ may include any measurable attributes of either
the choice maker or the park, and is typically represented as a linear-in-parameters
function of destination attributes $X_j$ and the travel cost $d_{ij}$,
\begin{equation}\label{eq:utilityV}
V_{ij} = X_j\beta + \beta_d * d_{ij}
\end{equation}
The coefficients $\vect{\beta}$ are frequently estimated from household
surveys, though in the absence of a survey we may assert reasonable values.

Note that the gravity formulation is itself a special case of the utility-based
specification where $\exp(X_j\beta + \beta_d d_{ij}) = X_j f(d_{ij})$ [@Daly1982].
The distinction between gravity and utility-based specification is meaningful, 
however. Primarily, it becomes possible to construct accessibility statistics based
on revealed behavioral preferences rather than calibrated or asserted values [@Handy1997].

Both gravity and utility-based specifications hold several advantages relative
to isochrone-based accessibility metrics more commonly found in the literature.
First, all individuals are defined as having some access to all parks, rather
than an arbitrary cutoff asserted by the researcher. This allows for the fact
that some people are more or less sensitive to distances, and that distance is a
continuous, non-binary phenomenon. It defies reason to assume a person living
1.1 miles from a park has functionally different accessibility than someone
living at 0.9 miles. Second, the random utility formulation allows the
researcher to include -- in principle -- any attribute of the park as part of
its utility specification. This suggests that not all parks are equal, and that
a large park with many amenities such as Central Park in Manhattan may provide
health and activity benefits over a much larger area than a smaller community
square.

@@

This study uses data available to the public from a variety of federal and state
data agencies. The datasets, as well as the analysis code, are available on GitHub
at \url{https://github.com/gregmacfarlane/parks_access}.

@@

The Centers for Disease Control and Prevention makes small-area estimates on
key health indicators available through its 500 Cities data program [@CDC5002016].
The indicators are multilevel aggregations and imputations of BRFSS responses
[@Wang2018; @Wang2017], and have been recently used to study the tract-level
link between gentrification and urban health [@Gibbons2018]. We use two
indicators as our dependent variables in this study: the share of adults
in a Census tract who are obese, and the share of adults who participate in no
leisure-time physical activity. To improve clarity in our interpretation, we use
the complement of the second variable --- the share of tract adults who
participate in *some* physical activity --- even if the amount is not sufficient
to affect overall health. Both indicators are estimated for the year 2016.

To the health data, we join socioeconomic data collected through the Census
Bureau API via the \texttt{tidycensus} package for R [@Walker2019]. The
primary dataset is a geographic polygons layer of Census tracts in the five
boroughs of New York City. We append to each Census tract relevant
sociodemographic data for each tract from the American Community Survey
2013-2017 5-year estimates. For a small handful of tracts in our sample, Census
supressed the median income estimate; these appear to be primarily wealthy
tracts and in almost all cases the CDC estimates of obesity and physical
activity are missing as well. After removing these tracts from the estimation
dataset, we have `r nrow(tracts)` complete cases. Table \ref{tab:tractsdata}
presents key descriptive statistics for these data.

@@

In a destination choice framework, the tracts represent the "origins" and the
"destinations" are parks and green spaces in New York City.
We retrieved a polygons layer of public parks and greenspaces within New York City’s
municipal boundaries and checked it for accuracy and relevance [@nycparks].
Upon inspection, we removed several facilities that do not qualify as
publicly accessible green space, such as Yankee Stadium, Citi Field, and their
surrounding parking lots. We also removed parks of less than half an acre in
size, as these appear to be predominantly planted medians rather than legitimate
public green space. We also consolidated individual geographic polygons
comprising a single facility --- as is the case in Flushing Meadows --- and
eliminated sub-facilities such as tennis courts or baseball fields included within
larger park facilities. Instead of these distinct sub-facilities, we created 
variables for each containing park indicating the presence of sports courts,
playgrounds, and trails.

To this dataset of parks and regularized green spaces, we add a polygons layer of
cemeteries in New York that are open to the public. Cemeteries are important
green spaces that can be used for many types of physical activity. These
operations leave us with `r nrow(parks)` distinct parks. For each park and cemetery
we generate calculate the size of the park in acres. Using a Twitter application
programming interface (API) implemented with the Python package Tweepy,  we also
collected and stored tweets containing geotags of precise geographic coordinates
located within the boundaries of the parks; from this information,
we were able to segregate tweets that were generated within a park in September
2014. This geolocated twitter activity could provide an additional point of information
on the degree to which individual parks are actually used [@Wang2016].

@@

We calculate a set of utility-based accessibility statistics for each tract
in New York City. The most basic utility specification includes only the
the park size in acres and the distance from the population-weighted tract centroid
to the boundary of the park in miles,
\begin{equation}\label{eq:u_access}
V_{ij} = \lambda_s * \log(size_j) + \lambda_d * \log(distance_{ij})
\end{equation}
The logarithmic transform allows for diminishing marginal utility of distance
and park size: A 1-mile increase to a trip matters more for a 1-mile trip than a
10-mile trip. @Macfarlane2019 estimated destination choice 
parameters for park trips in the Alameda County (Oakland), California using
this utility specification, obtaining values of $\lambda_s = 0.373$ and 
$\lambda_d = -1.98$. The ratio of these estimates implies people are willing to
travel roughly six times further to reach a park twice as large. In the absence 
of park trip destination choice coefficients in New York City, we apply these
previous estimates.

The second utility specification is adapted from a park destination choice model
estimated by @Kinnell2006 in New Jersey. In this study the authors used a revealed
preference survey to determine which factors of a park and its surrounding environment
affected its selection. By bringing over common variables, we can create a second
utility specification as
\begin{equation}\label{eq:u_multi}
V_{ij} = \lambda_s * \log(size_j) + \lambda_d * \log(distance_{ij}) + \lambda_t * trails + \\
  \lambda_{c} * courts + \lambda_p * playgrounds
\end{equation}
with $\lambda_t = 0.99$, $\lambda_c = 0.43$, and $\lambda_p = 0.26$. @Kinnell 
did not transform their size and distance estimates, so we retain the previously 
estimated and applied values. We assume that the other covariates in the @Kinnell
model which we do not have available to us (e.g., boat launches, picnic areas) are
orthogonal to the included parameters and leaving them out will not affect the
values of the included coefficients.

The final utility specification includes the number of geolocated tweets
emanating from within the park in September of 2014 in addition to the
basic specification in Equation \ref{eq:u_access},
\begin{equation}\label{eq:u_tweets}
V_{ij} = \lambda_s * \log(size_j) + \lambda_d * \log(distance_{ij}) + \lambda_{tw} * YJ(tweets)
\end{equation}
with $YJ(x)$ being a Yeo-Johnson transformation that implements a diminishing
marginal return similar to $log(x)$, but where $YJ(0) = 0$ [@Yeo2000]. In this
case we have no external information describing how destination choice to
parks may be affected by the twitter activity in the park, so we assert a 
value of $\lambda_{tw} = 0.1$

@@

We predict either the physical activity or the obesity rate ($y$) in a census
tract as a linear function of the tract's sociodemographic characteristics $X$
and accessibility to parks ($A$),
\begin{equation}\label{eq:themodel}
 y = X\beta + \beta_{a}A + \varepsilon
\end{equation}

@@

The specific spatial model applied is both an econometric choice
--- as improperly specified spatial models can lead to inconsistent estimates of
model parameters and standard errors --- as well as a philosophical choice about
the likely structure of spillovers in the problem at hand. In case of physical
activity and obesity rates, we believe that these outcomes
are locally determined: That is, a particular individual decides whether or not
to participate in physical activity independently of whether his or her neighbors
participate in physical activity. Mathematically, we assert that the autodependence
relationship on the dependent variable $y = \rho W y + \varepsilon$ has 
$\rho = 0$ in the current case.

With this relationship ruled out, it remains a possibility that the 
model residuals are spatially dependent,
\begin{equation}\label{eq:SEM}
 y = X\beta + \beta_a A + u; u = \lambda W u + \varepsilon
\end{equation}
that the outcome is partially dependent on the socioeconomic variables in *neighboring*
tracts,
\begin{equation}\label{eq:SLX}
 y = X\beta + \beta_a A + W X \lambda + \varepsilon
\end{equation}
or a linear combination of both.
\begin{equation}\label{eq:SDEM}
 y = X\beta + \beta_a A + W X \gamma + u; u = \lambda W u + \varepsilon 
\end{equation}
These models are referred as the spatial error model (SEM), the spatial lag of
$X$ model (SLX), and the spatial Durbin error model (SDEM). @Pace2008 suggest
that a Hausman-style test of the estimates of $\beta$ derived from the OLS 
(Equation \ref{eq:themodel}) and SEM (Equation \ref{eq:SEM}) can identify whether
the estimates are consistent. If the $\beta$ estimates are dissimilar, both 
specifications are inappropriate as they contain untreated omitted variables.
Similarly, if the SLX and SDEM estimates are dissimilar it is an indication of 
missing variables in that specification. On the other hand if the estimates of
both pairs of models are similar and the residual correlation parameter $\lambda$
is non-zero, the model accounting for spatially dependent residuals will produce 
proper estimates of the model standard errors.

>>>>

---
title: >
  Modeling the impacts of park access on health outcomes: a utility-based 
  accessibility approach
author:
  - name: Gregory Macfarlane
    email: gregmacfarlane@byu.edu
    affiliation: Brigham Young University
    corresponding: gregmacfarlane@byu.edu
  - name: Nico Boyd
    email:
    affiliation: Fehr and Peers 
  - name: John E. Taylor
    email: jetaylor@gatech.edu
    affiliation: Georgia Institute of Technology
  - name: Kari Watkins
    email: kariwatkins@gatech.edu
    affiliation: Georgia Institute of Technology
address:
  - code: Brigham Young University
    address: Civil and Environmental Engineering Department, Brigham Young University 430 Engineering Building, Provo, UT, 84602
  - code: Fehr and Peers
    address: Los Angeles, California
  - code: Georgia Institute of Technology
    address: School of Civil and Environmental Engineering, Georgia Institute of Technology, 755 Atlantic Drive, Atlanta, GA, 30332
abstract: >
  Recent research has underscored the potential for public green spaces to
  influence individual and societal health outcomes, but empirical measurements
  of such influences have yielded mixed results to date, with particular
  disagreement surrounding how access to parks ought to be defined while
  controlling for alternate explanations. In this paper we apply a comprehensive
  measure of park accessibility drawn from random utility choice theory, which
  avoids arbitrary assertions of proximity while incorporating potentially
  numerous amenities and attributes of both the parks and the population. We
  apply this utility-based accessibility measure to correlate Census tract-level
  obesity and physical activity rate estimates from the Centers for Disease
  Control and Prevention's 500 Cities project with tract-level American
  Community Survey socioeconomic data in New York City, paired with geographic
  open space data from New York City. Controlling for the
  socioeconomic variables and spatially correlated error terms, we show a
  positive and significant relationship between park access and physical
  activity rates. The data also suggest a negative relationship between park
  access and obesity rates, beyond what is expected through physical activity
  and socioeconomics.
  In doing so, this research contributes a more
  comprehensive modeling approach for measuring the impact of park access on
  health, and may improve our understanding of the role  parks and access to
  them can serve in furthering public health objectives.
author_summary: >
  Greg Macfarlane is an assistant professor of Civil Engineering at Brigham
  Young University in Provo, Utah.

@@

knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",", digits = 4, drop0trailing = TRUE)
})

@@

Historically, most large-scale health promotion efforts focused on
individual-level interventions intended to educate people about healthy
lifestyles and behaviors, touching on topics including diet and exercise.
Over the last several years a new *social* model of health has evolved,
described by @Duhl1999 
``as an outcome of the effects of socioeconomic status, culture,
environmental conditions, housing, employment and community influences''[-@Duhl1999, p.7].
In this new paradigm, resources provided via civil infrastructure --- in
particular parks and other public green spaces --- play a crucial role in
promoting and sustaining public health [@Bedimo-Rung2005; @Wells2007; @Coutts2008].
After all, it does little to lecture individuals on the importance of exercise
if their community is built in such a way that such exercise is expensive,
unenjoyable, or unsafe.

In spite of the potential for green spaces to influence public health outcomes 
and numerous studies attempting to empirically estimate the strength of these
outcomes, evidence to date has been mixed. A major challenge researchers have
faced is a proliferation of techniques for measuring access to green spaces, of
which many are plausible but lack robust theoretical basis. In this
paper, we apply a holistic and flexible measurement for park accessibility
based in activity location choice theory. This utility-based accessibility measure
compares the continuous distance to all parks in the region, weighted against
the sizes of each park and its other assorted amenities.

We apply this measurement to study the link between park accessibility and
attractiveness and Census tract-level aggregate physical activity participation
and obesity rates in New York City, controlling for spatially correlated
unobserved effects, spatial spillovers of covariates, and tract-level
socioeconomic characteristics. We also demonstrate how the utility-based
accessibility metric can potentially be expanded to account for other measures
of a park's attractiveness including its observed social media activity and the
presence of various park amenities. The paper concludes with a discussion of
opportunities for future research.

@@

The large majority of extant research on the topic of park access and public health
outcomes has been involved comparing metropolitan regions against each other. For example,
@West2012 used park data from
the Trust for Public Land’s 2010 "City Park Facts" and public health data from the
Behavioral Risk Factor Surveillance System (BRFSS) to examine the relationship
between the density of park land and physical activity and obesity rates for 67
metropolitan statistical areas in the U.S. The findings in this case conformed to 
expectation; that is, the study found a significant positive association
between park density and physical activity and a significant negative
association between park density and obesity.
@Larson2016 similarly used self-reported scores on the Gallup-Healthways
Wellbeing Index to evaluate the relationship between physical health and park
quantity, quality, and accessibility in 44 U.S. cities. While the authors found
positive associations between wellbeing and park quality and accessibility,
these relationships were not statistically significant.
Conversely, @Richardson2012 examined the relationship between total urban
greenspace and mortality rates for selected maladies; though the authors did not
find a statistically significant relationship between the quantity of urban
greenspace and mortality caused by lung cancer, diabetes, heart disease or car
accidents, they did find that all-cause mortality was --- oddly --- *higher* in greener
cities. Finally, in a meta-analysis of 20 peer reviewed journal articles exploring the
relationship between parks and objectively measured physical activity,
@Bancroft2015 found that five studies reported a significant positive
association between the two, six studies produced mixed results, and nine
studies found no association at all.

@@

Loosely defined, the *accessibility* of an arbitrary place describes the ease
with which people can accomplish activities there. Accessibility is an abstract
concept with tempting quantifiability [@Handy1997], perhaps explaining the proliferation of
calculation mechanisms. @Dong2006 present a helpful mathematical heirarchy of
common mechanisms, which we follow here. Consider a person residing at
point $i$ in a city with parks $j \in 1 \ldots J$. An analyst might consider
point $i$ as "having access" to a park if the distance to the park $d_{ij}$ is 
within an "isochrone," or a geometric buffer defined by some threshold $D$,

@@

Variations of this isochrone-based framework are easily derived and relatively
common. 
ParkScore [@parkscore2019] is a sophisticated application of this approach where $D$ is a
carefully calculated 10-minute network-bound walk, paired with a demographic analysis of
areas within and without this threshold. In other applications $D$ might be more
variably defined, such as the presence or percent of green space within a
political or statistical boundary [@Mitchell2008; @Stark2014] or within a specific distance
threshold [@Kaczynski2014]. It is also possible to attenuate an isochrone with 
various factors; for instance, @Dias2019 considered road safety perception as an
extenuating factor in the effective distance to a park.

In spite of the flexibility of adapted isochrone techniques, the arbritrary
definitions necessarily imposed by researchers in its application can inhibit
holistic analysis [@Logan2017].
A somewhat more complete approach is the so-called "gravity" accessibility
statistic. In this case the accessibility of point $i$ is the denominator of the
gravity formulation of a trip distribution model,
\begin{equation}\label{eq:gravity}
A_i = \sum_{j = 1}^J S_j f(d_{ij})
\end{equation}
where $S_j$ is the "size" of each park literally (i.e. acres) or abstractly (i.e.
trip attractions) and $f(d_{ij})$ is a monotonically decreasing cost function,
usually a negative exponential. @Dong2006 note that the isochrone framework is a
special case of the gravity model where $f(d_{ij})$ is a binary function and
$S_j = 1$.
@Giles-Corti2005 compare various gravity-based accessibility scores with an
isochrone specification and show that the former are more predictive of walking
behavior; that is, the attributes of a park matter influence walking more than
merely the distance to it. @Zhang2011 develop a population-weighted,
gravity-based accessibility to parks metric with a national scope, though they
do not examine its correlation with health outcomes.

A third accessibility mechanism is a utility-based specification, termed as such
for being derived from random utility choice theory. Consider that an individual
living at point $i$ is choosing a park for a recreation activity. If we apply
the multinomial logit model [@McFadden1974], the expected consumer surplus
enjoyed by this individual can be shown to be
\begin{equation}\label{eq:utility}
A_i = \ln\left({\sum_{j = 1}^J\exp(V_{ij})}\right) + C
\end{equation}
where parks are differentiated from each other by their relative measurable
*utilities*, $V_{ij}$. $C$ is an unknown constant, but the difference in consumer
surplus between two points $i$ and $k$ can be quantitatively compared as 
$A_i - A_k$ [@Bruce1977]. In principle, $V$ may include any measurable attributes of either
the choice maker or the park, and is typically represented as a linear-in-parameters
function of destination attributes $X_j$ and the travel cost $d_{ij}$,
\begin{equation}\label{eq:utilityV}
V_{ij} = X_j\beta + \beta_d * d_{ij}
\end{equation}
The coefficients $\vect{\beta}$ are frequently estimated from household
surveys, though in the absence of a survey we may assert reasonable values.

Note that the gravity formulation is itself a special case of the utility-based
specification where $\exp(X_j\beta + \beta_d d_{ij}) = X_j f(d_{ij})$
[@Daly1982]. The distinction between gravity and utility-based specification is
meaningful, however. Primarily, it becomes possible to construct accessibility
statistics based on revealed behavioral preferences rather than calibrated or
asserted values [@Handy1997]. @Kaczynski2016 present what is effectively a 
utility-based accessibility score they call "ParkIndex", with the explicit
motivation of developing a uniformly applicable park accessibility statistic,
though they limit this statistic as a way to add heterogeneity within an
isochrone analysis.

Both gravity and utility-based specifications hold several advantages relative
to isochrone-based accessibility metrics more commonly found in the literature.
First, all individuals are defined as having some access to all parks, rather
than an arbitrary cutoff asserted by the researcher. This allows for the fact
that some people are more or less sensitive to distances, and that distance is a
continuous, non-binary phenomenon. It defies reason to assume a person living
11 minutes via walking from a park always has meaningfully different
accessibility than someone living 9 minutes away. Second, the random utility
formulation allows the
researcher to include -- in principle -- any attribute of the park as part of
its utility specification. This suggests that not all parks are equal, and that
a large park with many amenities such as Central Park in Manhattan may provide
health and activity benefits over a much larger area than a smaller community
square.

@@

@@

This study uses data available to the public from a variety of federal and state
data agencies^[The datasets, as well as the analysis code, are available on GitHub
at \url{https://github.com/gregmacfarlane/parks_access}.].
The Centers for Disease Control and Prevention makes small-area estimates on
key health indicators available through its 500 Cities data program [@CDC5002016].
The indicators are multilevel aggregations and imputations of BRFSS responses
[@Wang2018; @Wang2017], and have been recently used to study the tract-level
link between gentrification and urban health [@Gibbons2018]. We use two
indicators as our dependent variables in this study: the share of adults
in a Census tract who are obese, and the share of adults who participate in no
leisure-time physical activity. To improve clarity in our interpretation, we use
the complement of the second variable --- the share of tract adults who
participate in *some* physical activity --- even if the amount may not be
sufficient to affect overall health. Both indicators are obtained for the year
2016.

To the health data, we join socioeconomic data collected through the Census
Bureau API via the \texttt{tidycensus} package for R [@Walker2019]. The
primary dataset is a geographic polygons layer of Census tracts in the five
boroughs of New York City. We append to each Census tract relevant
socioeconomic data for each tract from the American Community Survey
2013-2017 5-year estimates. For a small handful of tracts in our sample, Census
supressed the median income estimate; these appear to be primarily wealthy
tracts and in almost all cases the CDC estimates of obesity and physical
activity are missing as well. After removing these tracts from the estimation
dataset, we have `r nrow(tracts)` complete cases. Table \ref{tab:tractsdata}
presents key descriptive statistics for these data.

@@

In a destination choice framework, the tracts represent the "origins" and the
"destinations" are parks and green spaces in New York City.
We retrieved a polygons layer of public parks and greenspaces within New York City’s
municipal boundaries and checked it for accuracy and relevance [@nycparks].
Upon inspection, we removed several facilities that do not qualify as
publicly accessible green space, such as Yankee Stadium, Citi Field, and their
surrounding parking lots. We also removed parks of less than half an acre in
size, as these appear to be predominantly planted medians rather than legitimate
public green space. We consolidated individual geographic polygons
comprising a single facility --- as is the case in Flushing Meadows --- and
eliminated sub-facilities such as tennis courts or baseball fields included within
larger park facilities. Instead of these distinct sub-facilities, we created 
variables for each containing park indicating the presence of sports courts,
playgrounds, and trails.

To this dataset of parks and regularized green spaces, we add a polygons layer of
cemeteries in New York that are open to the public. Cemeteries are important
green spaces that can be used for many types of physical activity. These
operations leave us with `r nrow(parks)` distinct green spaces; for simplicity 
we will refer to both parks and cemeteries as "parks" going forward. 
For each park and 
we calculate the size of the park in acres. Using a Twitter application
programming interface (API) implemented with the Python package Tweepy,  we also
collected and stored tweets containing geotags of precise geographic coordinates
located within the boundaries of the parks; from this information,
we were able to segregate tweets that were generated within a park in September
2014. This geolocated twitter activity could provide an additional point of information
on the degree to which individual parks are actually used [@Wang2016].

@@

We calculate a set of utility-based accessibility statistics for each tract
in New York City. The most basic utility specification includes only the
the park size in acres and the distance from the population-weighted tract centroid
to the boundary of the park in miles,
\begin{equation}\label{eq:u_access}
V_{ij} = \lambda_s * \log(size_j) + \lambda_d * \log(distance_{ij})
\end{equation}
The logarithmic transform allows for diminishing marginal utility of distance
and park size: A 1-mile increase to a trip matters more for a 1-mile trip than a
10-mile trip. @Macfarlane2019 estimated destination choice 
parameters for park trips in the Alameda County (Oakland), California using
this utility specification, obtaining values of $\lambda_s = 0.373$ and 
$\lambda_d = -1.98$. The ratio of these estimates implies people are willing to
travel roughly six times further to reach a park twice as large. In the absence 
of park trip destination choice coefficients in New York City, we adopt these
previous estimates.

@Kinnell2006 surveyed park users in New Jersey and applied a multinomial logit
model to determine which factors influenced its perceived utility as a park
trip destination. By transferring estimates of common variables between our
data and this survey-based model, we can create a second utility specification
as
\begin{equation}\label{eq:u_multi}
V_{ij} = \lambda_s * \log(size_j) + \lambda_d * \log(distance_{ij}) + \lambda_t * trails + \\
  \lambda_{c} * courts + \lambda_p * playgrounds
\end{equation}
with $\lambda_t = 0.99$, $\lambda_c = 0.43$, and $\lambda_p = 0.26$. Kinnell et al.\
did not transform their size and distance estimates, so we retain the previously 
estimated and applied values. We assume that the other covariates in the 
model which we do not have available to us (e.g., boat launches, picnic areas) are
orthogonal to the included parameters and leaving them out will not affect the
values of the included coefficients.

The third utility specification includes the number of geolocated tweets
emanating from within the park in September of 2014 in addition to the
basic specification in Equation \ref{eq:u_access},
\begin{equation}\label{eq:u_tweets}
V_{ij} = \lambda_s * \log(size_j) + \lambda_d * \log(distance_{ij}) + \lambda_{tw} * YJ(tweets)
\end{equation}
with $YJ(x)$ being a Yeo-Johnson transformation that implements a diminishing
marginal return similar to $log(x)$, but where $YJ(0) = 0$ [@Yeo2000]. In this
case we have no external information describing how destination choice to
parks may be affected by the twitter activity in the park, so we assert a 
value of $\lambda_{tw} = 0.1$

@@

We predict either the physical activity or the obesity rate ($y$) in a census
tract as a linear function of the tract's sociodemographic characteristics $X$
and accessibility to parks $A$,
\begin{equation}\label{eq:themodel}
 y = X\beta + \beta_{a}A + \varepsilon
\end{equation}

@@

The specific spatial model applied is both an econometric choice
--- as improperly specified spatial models can lead to inconsistent estimates of
model parameters and standard errors --- as well as a philosophical choice about
the likely structure of spillovers in the problem at hand. In case of physical
activity and obesity rates, we believe that the outcomes are locally dependent:
That is, a particular individual decides whether or not to participate in
physical activity independently of whether his or her neighbors
participate in physical activity. This does not mean that we think it impossible that
the socioeconomic variables of neighboring tracts do not play a role. Mathematically, we assert that the autodependence
relationship on the dependent variable $y = \rho W y + \ldots$ has 
$\rho = 0$ in the current case.

With this relationship ruled out, it remains a possibility that the 
model residuals are spatially dependent,
\begin{equation}\label{eq:SEM}
 y = X\beta + \beta_a A + u; u = \lambda W u + \varepsilon
\end{equation}
that the outcome is partially dependent on the socioeconomic variables in *neighboring*
tracts,
\begin{equation}\label{eq:SLX}
 y = X\beta + \beta_a A + W X \gamma + \varepsilon
\end{equation}
or a linear combination of both.
\begin{equation}\label{eq:SDEM}
 y = X\beta + \beta_a A + W X \gamma + u; u = \lambda W u + \varepsilon 
\end{equation}
These models are referred as the spatial error model (SEM), the spatial lag of
$X$ model (SLX), and the spatial Durbin error model (SDEM). @Pace2008 suggest
that a Hausman-style test of the estimates of $\beta$ derived from the OLS 
(Equation \ref{eq:themodel}) and SEM (Equation \ref{eq:SEM}) can identify whether
the estimates are consistent. If the $\beta$ estimates are dissimilar, both 
specifications are inappropriate as they contain untreated omitted variables.
Similarly, if the SLX and SDEM estimates are dissimilar it is an indication of 
missing variables in that specification. On the other hand if the estimates of
both pairs of models are similar and the residual correlation parameter $\lambda$
is non-zero, the model accounting for spatially dependent residuals will produce 
proper estimates of the model standard errors.
